Trail Saving on Backtrack

### 2020-02-24 paper.tex
fb changes clean version (except for experiments)

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\usepackage[noend]{algcompatible}
%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{theorem}
\algrenewcommand\algorithmicprocedure{}
\algrenewcommand\algorithmicthen{}
\algrenewcommand\algorithmicdo{}

@@

\renewcommand{\note}[1]{{\color{red}{#1}}}
\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdxT}[2]{\ensuremath{\iota_{#1}(#2)}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\range}[2]{#1\ldots#2}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\reasonsave}[1]{\ensuremath{\mathit{reason_{\mathit{save}}(#1)}}}
\newcommand{\litsave}{\ensuremath{\mathit{l_{\mathit{save}}}}}
\newcommand{\formula}{\ensuremath{\mathcal{F}}}
\renewcommand{\implies}{\rightarrow}
\newcommand{\uipcls}{C_{\textit{1-UIP}}}
\newcommand{\deepestLvl}{L_{\textit{deep}}}
\newcommand{\deepestLit}{\ell_{\textit{deep}}}
\newcommand{\btL}{L_{\textit{back}}}
\newcommand{\cbt}{C-bt\xspace}
\newcommand{\trailsave}{\trail_{\mathit{save}}}

\newtheorem{Obs}{Observation}
\newtheorem{Cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{ex}{Example}
\newcommand{\whitebox}{\raisebox{.5ex}{\fbox{\hspace*{.2ex}}}}

\title{Trail Saving on Backtrack}
\author{Randy Hickey \and Fahiem Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada, 
  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}

\begin{document}
\maketitle
\begin{abstract}
    A CDCL \sat solver can backtrack a large distance when it learns a
    new clause, e.g, when the new learnt clause is a unit clause the
    solver has to backtrack to level zero. When the length of the
    backtrack is large the solver can end up reproducing many of the
    same decisions and propagations when it redescent the search
    tree. Motivated by this potential redundancy, partial backtracking
    (chronological backtracking) has been proposed. This technique
    allows the solver to reduce the length of its backtrack, e.g.,
    allowing the solver to backtrack to the previous level after a
    conflict is found. However, this technique has its shortcomings
    and only tends to work when limits are placed on its
    application. In this paper we present a new trail saving technique
    that makes a copy of the part of the trail that was backtracked
    over. This saved copy can then be used to improve the efficiency
    of the solver's subsequent redescent. Although our technique does
    not save as much work as chronological backtracking it also does
    not have chronological backtracking's shortcomings. Furthermore,
    the saved trail also provides the solver the ability to look ahead
    along the previous trail. This can be exploited to improve its
    efficiency. Our empirical results show that our technique often
    yields superior performance over chronological backtracking, and
    that is is able to improve the performance of state of the art
    solvers.

@@

\section{Introduction}
\textbf{Overall message of the paper.}
\begin{enumerate}
\item Formalism for understanding a more general form of trail savings
    which caches a copy of the trail that is backtracked over, and
    verifying how it exploited soundly.

\item Potential advantages over chronological backtracking.  Literals
    remain at the correct level, supports the solver having more
    freedom in its decisions.

\item Potential advantages of techique (1) increase in efficiency of
    propagation, (2) the new uses of the cached information.

\item Potential disadvantage. Reasons are recycled, in some circumstances
    new reasons might be better. We show in a simple experiment that
    this effect can occur.

\item Techique captures most of the positive effects of
    non-chronological backtracking in a simpler to implement way. It
    also preserves the simpler standard structure of CDCL
    solvers. Already provides some advantages for current SAT
    solvers. The ideas also have the potential for further
    exploitation. Lookahead for conflicts seems to be beneficial, and
    potentially if better discrimatory measures could be developed for
    reason clauses we could also better determine whether or not to
    use the cached information.. 
\end{enumerate}

\note{Reusing the Assignment Trail in CDCL Solvers: restart to level
  $i$ where the next decision has a higher score than the decision at
  level $i$ and a lower score than all decisions at level $j$ for
  $j< i$.

  They argum changes to reason clauses don't make much difference, but
  that was with minisat, before metrics for clause quality were in place in sat solvers.

  The savings for Luby-100 (used in minisat) were about 1\%}

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that backtracking chronologically
\cite{DBLP:conf/lpar/JiangZ13} (i.e. backtracking one level only after
conflict analysis)
\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
on many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

There are at least two possible side-effects of chronological
backtracking which may explain its effectiveness. It has been observed
that when a solver backtracks non-chronologically after a conflict,
many of the literals that are unassigned during the backtrack will be
re-assigned again in roughly the same order when the solver
continues. By using chronological backtracking, the solver keeps a lot
more of its partial assignment intact and saves from having to
reconstruct a lot of the trail via propagation that it otherwise would
have. Saving from having to reconstruct portions of the trail after
backtracking was first introduced in the context of restarts by van
der Tak et al. \cite{DBLP:journals/jsat/TakRH112}. The second
side-effect of chronological backtracking is that it forces the solver
to stay in the same local neighbourhood during search, since it
unassigns less of the partial assignment and one could argue that this
might lead to finding additional conflicts more quickly or reaching a
satisfying assignment (if one exists) more quickly. However, this
second side-effect also leads to less search flexibility, since the
solver does not jump out of local neighbourhoods as often and does not
obey the variable selection heuristic. We will argue that it is the
first side-effect that leads to the effectiveness of chronological
backtracking and not the second by developing a technique which also
saves from having to repeat propagations, but allows the search to
remain flexible and obey the variable selection heuristic.

Our alternative method is "trail saving", where we cache the part of
the trail that is unassigned during a backtrack and then attempt to
restore implied parts of it as each new decision is made. Trail saving
preserves the traditional invariants of the SAT solver and its basic
version is very simple to implement. It also allows the search to
choose the order of decisions, but helps make propagation faster. We
explore the theoretical speedup that trail saving provides, develop
some enhancements to make the idea more effective, and demonstrate
experimentally that it performs similarly well as chronological
backtracking without suffering from some of the aforementioned
drawbacks.

\section{Background}
A propositional formula $\formula$ expressed in Conjunctive Normal
Form (CNF), contains a set of variables $V$. A literal is a variable
$v\in V$ or its negation $\lnot v$, and for a literal $l$ we let
$\var(l)$ denote its underlying variable. A CNF consists of a
conjunction of clauses, each of which is a disjunction of literals. We
often view a clause as being a set of literals and employ set
notation, e.g., $\ell\in C$ and $C'\subset C$.

We assume the reader is familiar with the basic operations of CDCL
\sat solvers. A good source for this background is
\cite{DBLP:series/faia/SilvaLM09}.

\paragraph{Trails.}
CDCL \sat solvers maintain a trail which is the sequence of literals
that have currently been assigned \true by the solver. During its
operation a \sat solver will add newly assigned literals to the end of
the trail, and on backtrack remove literals from the end of the trail
(unassigning them as they are removed).

A \sat solver's trail satisfies a number of conditions that we will
discuss below. However, in we will need some additional flexibility in
our definitions, as we will be sometimes be working with trails that
would never be constructed by a \sat solver. Hence, we define a
\emph{trail} to be a sequence of literals each of which is either a
\emph{decision} literal or an \emph{implied} literal, and each of
which has a \emph{reason}. These two types of literals are
distinguished by their \emph{reasons}. Decision literals $d$ have a
null reason, $\reason{d} = \varnothing$. Implied literals $l$ have as
a reason a clause of the formula $\formula$,
$\reason{l} = C\in\formula$. ($\reason{l}$ can be a learnt clause that
has been added to $\formula$).

If literal $\ell$ is on the trail $\trail$ let
$\trailIdxT{\trail}{\ell}$ denote its index on the trail, i.e,
$\trail[\trailIdxT{\trail}{\ell}] = \ell$. For convenience, we also let
$\trailIdxT{\trail}{\ell} = \trailIdxT{\trail}{\lnot \ell} =
\trailIdxT{\trail}{\var(\ell)}$ even though neither $\lnot \ell$ nor
$\var(\ell)$ are actually on $\trail$. If $x$ and $y$ are both on the
trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say that
\textit{$x$ appears before $y$ on the trail}. For convenience, when
the trail being discussed in clear from context we simply write
$\trailIdx{}$ instead of $\trailIdxT{\trail}{}$.

Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
which is equal to the number of decision literals appearing before
$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$. The set of literals on $\trail$ that
have the same decision level forms a \emph{contiguous}
subsequence\footnote{Our approach uses standard trails in which the
  decision levels are contiguous. Chronological backtracking
  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
  generates trails with non-contigous decision levels} that starts
with a decision literal $d_i$ and ends just before the next decision
literal $d_{i+1}$. We will often need to refer to different decision
level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
subsequence of literals at decision level $i$; and let
$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
decision levels $k$ for $i\leq k\leq j$. Now we define the
following properties that a trail $\trail$ can have.
\begin{description}
\item[non-contradictory:] A variable cannot appear in both polarities
    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
\item[non-redundant:] A literal can only appear once on $\trail$.
\item[reason-sound:] For each implied literal $l\in\trail$ we have
    that it reason clause has \emph{been made unit by $\trail$
      implying $l$}. This means that the reason clause contains $l$
    and the negation all of its other literals are on $\trail$ before
    $l$:
    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
    \trailIdx{l}\bigr)$.
\item[propagation-complete:] Unit propgation has been run to
    completion at all decisions levels of $\trail$. This means that
    literals appear on $\trail$ at the first decision level they were
    unit implied. Formally, this can be captured by the condition:
    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. With
    non-redundancy $l$ will appear only at the first decision level it
    was implied, without non-redundancy, $l$ might appear more than
    once. Note that this also implies that $\reason{l}$ must contain
    at least one other literal $y\neq l$ with
    $\dlevel{y} = \dlevel{l}$.
\item[conflict free:] No clause of $F$ is falsified by
    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
    called \emph{conflicts}.
\end{description}

@@

In standard backtracking, the solver then constructs a new 1-UIP
clause by resolving away all but one literal at level $\deepestLvl$
from the conflict $C$ using the reason clauses of these literals. (As
noted above $C$ must contain at least two literals at level
$\deepestLvl$). Hence, the new clause $\uipcls$ will contain one
literal $\deepestLit$ at level $\deepestLvl$ and have all of its other
literals a levels less than $\deepestLvl$. The solver then backtracks
to $\btL$ the second deepest level in $\uipcls$. This involves
changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$
(unassigning all literals removed from $\trail$). The new clause
$\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$ implying
$\deepestLit$, so the solver then adds $\deepestLit$ to the trail and
executes another round of unit propagation at level $\btL$, after
which it continues by once again growing the trail with new decisions
and unit implied literals until a new conflict or a satisfying
assignment is found.

@@

In \cite{DBLP:journals/jsat/TakRH11} a technique for reducing the
length of the backtrack during restarts was presented. In restarts,
the solver backtracks to level $0$, and this technique involves
computing a new deeper backtrack level $M> 0$ for which it is known
that on redescent the first $M+1$ levels of the trail will be
unchanged (except perhaps for the ordering of the literals). This
technique removes the redundant work of reproducing the first $M$
trail levels. Unfortunately, with backtracking the trail will be
changed at level $\btL$ ($\deepestLit$ will be newly inserted at this
level). Thus there seems to be no choice but to backtrack to
$\btL$. In this paper we will show that although we have to backtrack
to $\btL$ we can make the subsequent redescent much more efficient.

\paragraph{Chronological Backtracking.}
Chronological bactracking (\cbt) is a technique originally propsed by
Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
\textit{partial backtracking}. Its aim is to reduce the redundant work
that might be done by the \sat solver on its redescent from the
backtrack level $\btL$. The original paper proposed a technique for
backtracking to any level $j$ in the range
$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
the conflict was discovered. Nadel and Ryvchin
\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
flexibly backtracking any level in the allowed range. Note that the
new 1-UIP clause $\uipcls$ is made unit at every level in this
range. So after backtracking to level $j$ the newly implied literal
$\deepestLit$ is added to the trail with
$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
$\btL$ (the second deepest level in $\uipcls$.

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences on the \sat solvers operation, all of which were
described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
\cite{DBLP:conf/sat/NadelR18} some more practical implementation
methods were developed to overcome this issues, and in
\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
developed under which the changes required to the \sat solver could be
shown to be successful. We refer the reader to the cited papers for
more details.

@@

The reduction in redundant work achieved by \cbt does not come for
free. In particular, some of the work avoided by \cbt is not actually
redundant work; it is work that can actually help the solver. The
obvious evidence for this statement is that in both
\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
found that it was not optimal to always apply \cbt. In
\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
the standard backtrack, $\btL-\deepestLvl$ was greater than a
given threshold $T$. In their experiments they found that $T=100$ was
the best value, i.e., \cbt is done only on longer backtracks. In
practice, this meant that \cbt was \emph{hardly ever done}; in our
measurements with their solver only about 3\% of the solver backtracks
were \cbt backtracks. In \cite{DBLP:conf/sat/MohleB19} the value
$T=100$ was also applied. However, they introduced an additional
technique to add some applications of \cbt when the length of the
backtrack is less than $T$. This allowed to utilize \cbt in \%15 of
the backtracks. Nevertheless, the fact \cbt has to be used in a
controlled fashion, indicates that can be benefits in the \sat solver
employing standard backtracking; it cannot be the case that every
redescent is only doing redundant work.

>>>>

\documentclass[runningheads]{llncs}
\usepackage{times}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\usepackage[noend]{algcompatible}
%\usepackage[ruled,linesnumbered,noend,oldcommands]{algorithm2e}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}
\usepackage[table]{xcolor}
\usepackage{theorem}

\algrenewcommand\algorithmicprocedure{}
\algrenewcommand\algorithmicthen{}
\algrenewcommand\algorithmicdo{}
\definecolor{linen}{rgb}{0.98, 0.94, 0.9}
\definecolor{lightcyan}{rgb}{0.88, 1.0, 1.0}

@@

\renewcommand{\note}[1]{{\color{red}{#1}}}
\newcommand{\fs}[1]{\fontsize{#1}{#1}\selectfont}
\newcommand{\fss}[2]{\fontsize{#1}{#2}\selectfont}
\newcommand{\sat}{SAT\xspace}
\newcommand{\trail}{\ensuremath{\mathcal{T}}}
\newcommand{\trailIdxT}[2]{\ensuremath{\iota_{#1}(#2)}}
\newcommand{\trailIdx}[1]{\ensuremath{\iota(#1)}}
\newcommand{\range}[2]{#1\ldots#2}
\newcommand{\dlevel}[1]{\ensuremath{\mathit{decLvl}(#1)}}
\newcommand{\dlevels}{\ensuremath{\mathit{decLvls}}}
\newcommand{\var}{\text{var}}
\newcommand{\true}{\textsc{true}\xspace}
\newcommand{\false}{\textsc{false}\xspace}
\newcommand{\reason}[1]{\ensuremath{\mathit{reason}(#1)}}
\newcommand{\reasonsave}[1]{\ensuremath{\mathit{reason_{\mathit{save}}(#1)}}}
\newcommand{\litsave}{\ensuremath{\mathit{l_{\mathit{save}}}}}
\newcommand{\formula}{\ensuremath{\mathcal{F}}}
\renewcommand{\implies}{\rightarrow}
\newcommand{\uipcls}{C_{\textit{1-UIP}}}
\newcommand{\deepestLvl}{L_{\textit{deep}}}
\newcommand{\deepestLit}{\ell_{\textit{deep}}}
\newcommand{\btL}{L_{\textit{back}}}
\newcommand{\cbt}{C-bt\xspace}
\newcommand{\trailsave}{\trail_{\mathit{save}}}
\newcommand{\bt}{\textsc{backtrack}\xspace}
\newcommand{\ust}{\textsc{useSavedTrail}\xspace}

\newtheorem{Obs}{Observation}
\newtheorem{Cor}{Corollary}
\newtheorem{defn}{Definition}
\newtheorem{thm}{Theorem}
\newtheorem{prop}{Proposition}
\newtheorem{ex}{Example}
\newcommand{\whitebox}{\raisebox{.5ex}{\fbox{\hspace*{.2ex}}}}

\title{Trail Saving on Backtrack}
\author{Randy Hickey \and Fahiem Bacchus}
\institute{Department of Computer Science, University of Toronto, Canada\\
  \email{rgh000@gmail.com, fbacchus@cs.toronto.edu}}

\begin{document}
\maketitle
\begin{abstract}
    A CDCL \sat solver can backtrack a large distance when it learns a
    new clause, e.g, when the new learnt clause is a unit clause the
    solver has to backtrack to level zero. When the length of the
    backtrack is large the solver can end up reproducing many of the
    same decisions and propagations when it redescent the search
    tree. Motivated by this potential redundancy, partial backtracking
    (chronological backtracking) has been proposed. This technique
    allows the solver to reduce the length of its backtrack, e.g.,
    allowing the solver to backtrack to the previous level after a
    conflict is found. However, this technique has its shortcomings
    and only tends to work when limits are placed on its
    application. In this paper we present a new trail saving technique
    that makes a copy of the part of the trail that was backtracked
    over. This saved copy can then be used to improve the efficiency
    of the solver's subsequent redescent. Although our technique does
    not save as much work as chronological backtracking it also does
    not have chronological backtracking's shortcomings. Furthermore,
    the saved trail also provides the solver with the ability to look
    ahead along the previous trail. This can be exploited to improve
    its efficiency. Our empirical results show that our technique
    often yields superior performance over chronological backtracking,
    and that is is able to improve the performance of state of the art
    solvers.

@@

\section{Introduction}
\textbf{Overall message of the paper.}

% \begin{enumerate}
% \item Formalism for understanding a more general form of trail savings
%     which caches a copy of the trail that is backtracked over, and
%     verifying how it exploited soundly.

% \item Potential advantages over chronological backtracking.  Literals
%     remain at the correct level, supports the solver having more
%     freedom in its decisions.

% \item Potential advantages of technique (1) increase in efficiency of
%     propagation, (2) the new uses of the cached information.

% \item Potential disadvantage. Reasons are recycled, in some circumstances
%     new reasons might be better. We show in a simple experiment that
%     this effect can occur.

% \item Technique captures most of the positive effects of
%     non-chronological backtracking in a simpler to implement way. It
%     also preserves the simpler standard structure of CDCL
%     solvers. Already provides some advantages for current SAT
%     solvers. The ideas also have the potential for further
%     exploitation. Lookahead for conflicts seems to be beneficial, and
%     potentially if better discrimatory measures could be developed for
%     reason clauses we could also better determine whether or not to
%     use the cached information.. 
% \end{enumerate}

% \note{Reusing the Assignment Trail in CDCL Solvers: restart to level
%   $i$ where the next decision has a higher score than the decision at
%   level $i$ and a lower score than all decisions at level $j$ for
%   $j< i$.

%   They argum changes to reason clauses don't make much difference, but
%   that was with minisat, before metrics for clause quality were in place in sat solvers.

%   The savings for Luby-100 (used in minisat) were about 1\%}

The vast majority of modern SAT solvers that are used to solve
real-world problems are based on the conflict-driven clause learning
(CDCL) algorithm. In a CDCL SAT solver, backtracking occurs after
every conflict, where all literals from one or more decision levels
become unassigned, before the solver resumes making decisions and
performing unit propagations. Traditionally, CDCL solvers would
backtrack non-chronologically to the conflict level, which is the
second highest decision level remaining in the conflict clause after
conflict analysis has resolved away all but one literal from the
current decision level \cite{DBLP:conf/dac/MoskewiczMZZM01}. Recently,
however, it has been shown that partial backtracking
\cite{DBLP:conf/lpar/JiangZ13} or chronological backtracking, \cbt,
(i.e. backtracking one level only after conflict analysis)
\cite{DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19} can be effective
on many instances and was implemented in solvers that won the last two
SAT competitions. Although chronological backtracking breaks some of
the conventional invariants of CDCL solvers, it has been formalized
and proven correct \cite{DBLP:conf/sat/MohleB19}.

The motivation for using \cbt is the observation that when a solver
backtracks across many levels, many of the literals that are
unassigned during the backtrack will be re-assigned again in roughly
the same order when the solver redescends. This observation was first
made in the context of restarts by van der Tak et
al. \cite{DBLP:journals/jsat/TakRH11}. Their technique backtracks to
the minimum change level, i.e., the first level at which the solver's
trail can change on redescent. However, their technique cannot be used
when backtracking from a conflict: the solver's trail is going to be
changed at the backtrack level so the minimum change level is the same
as the backtrack level.

Chronological backtracking or partial backtracking instead allows a
reduction in the length of the backtrack by placing literals on the
trail out of decision level order. By reducing the length of the
backtrack the solver can keep more of its assignment trail
intact. This can save it from the work involved in reconstructing a
lot of its trail. Using \cbt is not a panacea however. Its application
must be limited for peak effectiveness. This indicates that it is
sometimes beneficial for the solver backtrack fully and redo its
trail, even if this takes more work. We will expand on why this might
be case below.

In this paper we present an alternative ``trail savings'' method where
by we save the backtracked part of the solver's trail and attempt to
use that information to make the solver's redescent more
efficient. Trail saving preserves the traditional invariants of the
SAT solver and its basic version is very simple to implement. It
allows the search to retain complete control over the order of
decisions, but helps make propagation faster. We develop some
enhancements to make the idea more effective, and demonstrate
experimentally that it performs as well as and sometimes better than
chronological backtracking. We also show that with the enhancements we
are able to improve the performance of state of the art solvers.

\section{Background}
\label{sec:background}
\sat solvers determine the satisfiability of a propositional formula
$\formula$ expressed in Conjunctive Normal Form (CNF). $\formula$
contains a set of variables $V$. A literal is a variable $v\in V$ or
its negation $\lnot v$, and for a literal $l$ we let $\var(l)$ denote
its underlying variable. A CNF consists of a conjunction of clauses,
each of which is a disjunction of literals. We often view a clause as
being a set of literals and employ set notation, e.g., $\ell\in C$ and
$C'\subset C$. We will assume that the reader is familiar with the
basic operations of CDCL \sat solvers. A good source for this
background is \cite{DBLP:series/faia/SilvaLM09}.

\paragraph{Trails.}
CDCL \sat solvers maintain a trail which is the sequence of literals
that have currently been assigned \true by the solver. During its
operation a \sat solver will add newly assigned literals to the end of
the trail, and on backtrack remove literals from the end of the
trail. For convenience, we will regard \emph{literals as having been
  assigned \true if and only if they are on the trail}. So
removing/adding a literal to the trail is equivalent to
unassigning/assigning the literal $\true$.

A \sat solver's trail satisfies a number of conditions. However, in
this work we will need some additional flexibility in our definitions,
as we will be sometimes be working with trails that would never be
constructed by a \sat solver. Hence, we define a \emph{trail} to be a
sequence of literals each of which is either a \emph{decision} literal
or an \emph{implied} literal, and each of which has a
\emph{reason}. These two types of literals are distinguished by their
\emph{reasons}. Decision literals $d$ have a null reason,
$\reason{d} = \varnothing$. Implied literals $l$ have as a reason a
clause of the formula $\formula$, $\reason{l} =
C\in\formula$. (The clause $\reason{l}$ can be a learnt clause that has been
added to $\formula$).

If literal $\ell$ is on the trail $\trail$ let
$\trailIdxT{\trail}{\ell}$ denote its index on the trail, i.e,
$\trail[\trailIdxT{\trail}{\ell}] = \ell$. If $x$ and $y$ are both on
the trail and $\trailIdxT{\trail}{x} < \trailIdxT{\trail}{y}$ we say
that \textit{$x$ appears before $y$ on the trail}. For convenience,
when the trail being discussed is clear from context we simply write
$\iota$ instead of $\iota_{\trail}$.

Each literal $\ell\in\trail$ has a decision level $\dlevel{\ell}$
which is equal to the number of decision literals appearing before
$\ell$ on the trail plus one; hence, $\dlevel{d}=1$ for the first
decision literal $d\in\trail$. The set of literals on $\trail$ that
have the same decision level forms a \emph{contiguous}
subsequence\footnote{Our approach uses standard trails in which the
  decision levels are contiguous. Chronological backtracking
  \cite{DBLP:conf/lpar/JiangZ13,DBLP:conf/sat/NadelR18,DBLP:conf/sat/MohleB19}
  generates trails with non-contiguous decision levels} that starts
with a decision literal $d_i$ and ends just before the next decision
literal $d_{i+1}$. We will often need to refer to different decision
level subsequences of $\trail$. Hence, we let $\trail[[i]]$ denote the
subsequence of literals at decision level $i$; and let
$\trail[[\range{i}{j}]]$ denote the subsequence of literals at
decision levels $k$ for $i\leq k\leq j$. 
\begin{defn}
    A clause $C$ has \textbf{been made unit by $\trail$ implying $l$}
    when
    $l\in C \land \bigl(\forall x\in C. x\neq l \implies \lnot x \in
    \trail\bigr)$. That is, all literals in $C$ except $l$ must have
    been falsified by $\trail$
\end{defn}

Now we define the following properties that a trail $\trail$ can have.
\begin{description}
\item[non-contradictory:] A variable cannot appear in both polarities
    in the trail: $l\in \trail\implies \lnot l \not \in \trail$.
\item[non-redundant:] A literal can only appear once on $\trail$.
\item[reason-sound:] For each implied literal $l\in\trail$ we have
    that its reason clause $\reason{l}=C$ has been made unit by
    $\trail$ implying $l$, and for each $x\in C$ with $x\neq l$ we
    that $\lnot x$ appears before $l$ on $\trail$:
    $\forall l\in\trail.\, \reason{l}\neq \varnothing \implies
    l\in\reason{l} \land \bigl(\forall x\in\reason{l}.\, x\neq l
    \implies \lnot x \in \trail \land \trailIdx{\lnot x} <
    \trailIdx{l}\bigr)$.
\item[propagation-complete:] Unit propagation has been run to
    completion at all decisions levels of $\trail$. This means that
    literals appear on $\trail$ at the first decision level they were
    unit implied. Formally, this can be captured by the condition:
    $\forall i \in \{\dlevel{l}\,|\,l\in \trail\}. \bigl(\exists
    C\in\formula. \mbox{$C$ is made unit by $\trail[[\range{0}{i}]]$
      implying $l$}\bigr) \implies l\in \trail[[\range{0}{i}]]$. Note
    that propagation completeness implies that
    $\reason{l}\neq \varnothing$ must contain at least one other
    literal $y\neq l$ with $\dlevel{y} = \dlevel{l}$.
\item[conflict free:] No clause of $F$ is falsified by
    $\trail$. Clauses $C\in F$ falsified by $\trail$ are typically
    called \emph{conflicts}.
\end{description}

@@

In standard backtracking, the solver then constructs a new 1-UIP
clause by resolving away all but one literal at level $\deepestLvl$
from the conflict $C$ using the reason clauses of these literals. (As
noted above $C$ must contain at least two literals at level
$\deepestLvl$). Hence, the new clause $\uipcls$ will contain one
literal $\deepestLit$ at level $\deepestLvl$ and have all of its other
literals a levels less than $\deepestLvl$. The solver then backtracks
to $\btL$ the second deepest level in $\uipcls$. This involves
changing $\trail$ to its prefix $\trail[[\range{0}{\btL}]]$ (by our
convention all literals removed from $\trail$ are now unassigned). The
new clause $\uipcls$ is made unit by $\trail[[\range{0}{\btL}]]$
implying $\deepestLit$, so the solver then adds $\deepestLit$ to the
trail and executes another round of unit propagation at level $\btL$,
after which it continues by once again growing the trail with new
decisions and unit implied literals until a new conflict or a
satisfying assignment is found.

@@

In \cite{DBLP:journals/jsat/TakRH11} a technique for reducing the
length of the backtrack during restarts was presented. In restarts,
the solver backtracks to level $0$, and this technique involves
computing a new deeper backtrack level $M> 0$ for which it is known
that on redescent the first $M+1$ levels of the trail will be
unchanged (except perhaps for the ordering of the literals). This
technique removes the redundant work of reproducing the first $M$
trail levels. When backtracking from a conflict, however, the trail
will be changed at level $\btL$ ($\deepestLit$ will be newly inserted
at this level). Hence this technique cannot reduce the length of the
backtrack. In this paper we will show that although we have to
backtrack to $\btL$ we can make the subsequent redescent much more
efficient.

\paragraph{Chronological Backtracking.}
Chronological bactracking (\cbt) is a technique originally proposed by
Jiang and Zhang \cite{DBLP:conf/lpar/JiangZ13} under the name
\textit{partial backtracking}. Its aim is to reduce the redundant work
that might be done by the \sat solver on its redescent from the
backtrack level $\btL$. The original paper proposed a technique for
backtracking to any level $j$ in the range
$\btL \leq j \leq \deepestLvl{-}1$ (where $\deepestLvl$ is the level
the conflict was discovered). Nadel and Ryvchin
\cite{DBLP:conf/sat/NadelR18} proposed to always backtrack to
$\deepestLvl{-}1$ while M{\"{o}}hle and Biere
\cite{DBLP:conf/sat/MohleB19} returned to the original proposal of
flexibly backtracking to any level in the allowed range. Note that the
new 1-UIP clause $\uipcls$ is made unit at every level in this
range. So after backtracking to level $j$ the newly implied literal
$\deepestLit$ can be added to the trail with
$\reason{\deepestLit}=\uipcls$, and $\dlevel{\deepestLit}$ is set to
$\btL$ (the second deepest level in $\uipcls$.

This means that the decision levels on the trail are no longer
contiguous, as $\deepestLit$ has a different level than the other
literals at level $j$ (if $j\neq \btL$). This change has a number of
consequences for the \sat solver's operation, all of which were
described in the original paper \cite{DBLP:conf/lpar/JiangZ13}. In
\cite{DBLP:conf/sat/NadelR18} some more practical implementation
methods were developed to overcome this issues, and in
\cite{DBLP:conf/sat/MohleB19} a formal framework for \cbt was
developed under which the changes required to the \sat solver could be
shown to be sound. We refer the reader to the cited papers for
more details.

@@

The reduction in redundant work achieved by \cbt does not come for
free. In particular, some of the work avoided by \cbt is not actually
redundant work; it is work that can actually help the solver. The
obvious evidence for this statement is that in both
\cite{DBLP:conf/sat/NadelR18} and \cite{DBLP:conf/sat/MohleB19} it was
found that it was not optimal to always apply \cbt. In
\cite{DBLP:conf/sat/NadelR18} \cbt was applied only when the length of
the standard backtrack, $\btL-\deepestLvl$ was greater than a given
threshold $T$. In their experiments they found that $T=100$ was the
best value, i.e., \cbt is done only on longer backtracks. In practice,
this meant that \cbt was \emph{hardly ever done}; in our measurements
with their solver only about 3\% of the solver backtracks were \cbt
backtracks. In \cite{DBLP:conf/sat/MohleB19} the value $T=100$ was
also applied. However, they introduced an additional technique to add
some applications of \cbt when the length of the backtrack is less
than $T$. This allowed \cite{DBLP:conf/sat/MohleB19} to utilize \cbt
in about \%15 of the backtracks. Nevertheless, the fact \cbt has to be used
in a controlled fashion, indicates that there can be benefits in the \sat
solver employing standard backtracking; it cannot be the case that
every redescent is only doing redundant work.
